---
title: "0_preprocess_evaluationdataset"
author: "tsuyoshi"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: "hide"
    pandoc_args: [ 
      "--output=0_preprocess_evaluationdataset.html" 
    ]
date: "`r format(Sys.time(), '%d %B, %Y')`"

---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(imcRtools)
library(zellkonverter)
library(SingleCellExperiment)
library(cytomapper)

library(readr)
library(stringr)
library(tidyr)
library(dplyr)
library(common)

```


```{r specify_paths}

project.path <- "/mnt/projects/data/0_preprocess_th186"
# Specify input paths
steinbock.paths <- dir.find(path = file.path(project.path,"steinbock"), pattern = "th186_*")
  

meta.paths <- file.path(steinbock.paths, "images.csv")
images.paths <- file.path(steinbock.paths, "img")
masks.paths <- file.path(steinbock.paths, "masks")

# path to phi
phi.path <- file.path(project.path,"steinbock","phi.csv")
markergroup.path <-   file.path(project.path,"steinbock","marker_group.csv")
# Specify output path
out.path <- file.path(project.path,"processed")
# Create output directory if id doesn't exist
if (!dir.exists(out.path)) {dir.create(out.path, recursive=TRUE)}
```


## read sce and metadata files
all mcd files were merged into one sce object
```{r read_input_data}
sces <- lapply(steinbock.paths, function(x){read_steinbock(x, graphs_folder=NULL,regionprops_folder=NULL)} ) 
metas <- lapply(meta.paths, function(x){read_csv(x, show_col_types = FALSE)} ) 

sce <- do.call("cbind",sces)
meta <- do.call("rbind",metas)
```

## read panel for rowdata
```{r}
# path to panel
panel.path <- file.path(project.path,"steinbock","panel.csv")
panel <- read_csv(panel.path,show_col_types = F)

#set rownames
rownames(sce) <- panel$name[match(rowData(sce)$channel, panel$channel)] 
rowData(sce)
```


## add_tissue_annotation
```{r add_tissue_annotation}
sce$ROI_fullname <- meta$recovery_file[match(paste0(sce$sample_id,".tiff"), meta$image)] %>%
  str_split('.txt') %>% lapply('[[', 1) %>% unlist()
# add slide name
sce$slide <- sce$ROI_fullname %>% str_split('_') %>% lapply('[[', 4) %>% unlist()
# add tissue name 
sce$tissue <- sce$ROI_fullname %>% str_split('_') %>% lapply('[[', 5) %>% unlist()

# add ROI_per_tissue
sce$ROI_per_tissue <- sce$ROI_fullname %>% str_split('_') %>% lapply('[[', 6) %>% unlist()


table(sce$slide, sce$tissue)
```


## remove nuclei channel for further analyses
```{r remove_nucleichannel}
sce <- sce[!rownames(sce)%in%c("191Ir","193Ir","HH3"),]
```

## add "GT" or "CISI" channel for separation in following analyses
```{r}
rowData(sce)$channel_type <- 
  rownames(sce) %>% str_split('_') %>% lapply('[[', 1) %>% unlist()
```

## save unfiltered SCE
```{r}
writeH5AD(SingleCellExperiment(list(counts=assay(sce, "counts")),
                                 colData=colData(sce),
                                 rowData=rowData(sce)), file=file.path(out.path, "sce.h5ad"))
```

## Filter SCE
CD3 channel contained noticable unspecific signals.

CD3 outliers and Other bright specles from all channel were removed by manual thresholding.
```{r}
# Define CD3 outliers (Bright speckles)
CD3_thd <- 6.5
CD3_outliers <- as.vector(counts(sce["GT_CD3",])) >= CD3_thd

# Define CISI outliers (speckles)
CISI_thd <- 10
CISI_maxcounts <- sce[rowData(sce)$channel_type == "CISI",] %>% counts() %>% as.matrix() %>% apply( 2, max)
CISI_outliers <- CISI_maxcounts >= mean(CISI_maxcounts) + CISI_thd*sd(CISI_maxcounts)

# Define CIGTSI outliers (speckles)
GT_thd <- 10
GT_maxcounts <- sce[rowData(sce)$channel_type == "GT",] %>% counts() %>% as.matrix() %>% apply( 2, max)
GT_outliers <- GT_maxcounts >= mean(GT_maxcounts) + GT_thd*sd(GT_maxcounts)
sce[,GT_outliers]
# combibne outliers
outliers <- CD3_outliers | CISI_outliers | GT_outliers

# Filter sce
filtered_sce <- sce
counts(filtered_sce[,outliers]) <- counts(filtered_sce[,outliers]) *0

```


## save filtered SCE
```{r save_sce_as_anndata}
writeH5AD(SingleCellExperiment(list(counts=assay(filtered_sce, "counts")),
                                 colData=colData(filtered_sce),
                                 rowData=rowData(filtered_sce)), 
          file=file.path(out.path, "filtered_sce.h5ad"))

```





