---
title: "2_SMAF_parameters"
author: "tsuyoshi"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: "hide"
    pandoc_args: [ 
      "--output=2_SMAF_parameters.html" 
    ]
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r library}
library(ggplot2)
library(cowplot)
library(dplyr)
library(metan)
library(tidyr)
library(lessR)
```


### set up
```{r specify_paths}
# Set input path to the result
input_path <- "/mnt/projects/data/publication/2_SMAF_parameters"
U_path <- file.path(input_path, "U")
```

#### Read summary tables
```{r read_df}
fn <- file.path(input_path,"result.csv")
df <- read.csv(fn, header = T)

col_names <- c("methodW","ldaU","num_blocks_W","ldaW","k")
df[,col_names] <- lapply(df[,col_names] , factor)
df

```

### General analysis

#### 1. OMP_fixedk
```{r plot_results1, fig.height = 3, fig.width = 10, fig.align = "center"}

cur_df <- df[df$methodW=="omp_fixedk",]
p1 <- ggplot(cur_df, aes(x = ldaU, y = U_l0_mean, colour = k)) + geom_point(size = 1) +
  theme_bw() + ylim(0,max(cur_df$U_l0_mean)) + theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- ggplot(cur_df, aes(x = ldaU, y = d_modules, colour = k)) + geom_point(size = 1)  +
  theme_bw() + ylim(0,max(cur_df$d_modules)) + theme(axis.text.x = element_text(angle = 45, hjust = 1))

p3 <- ggplot(cur_df, aes(x = ldaU, y = SMAF_W_l0_mean, colour = k)) + geom_point(size = 1) +
  theme_bw() + ylim(0,max(cur_df$SMAF_W_l0_mean)) + theme(axis.text.x = element_text(angle = 45, hjust = 1))

p4 <- ggplot(cur_df, aes(x = ldaU, y = SMAF_Fit, colour = k)) + geom_point(size = 1)  +
  theme_bw() + ylim(0,1) + theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot_grid(p1,p2,p3,p4, nrow=1)
```



#### 2. Lasso
```{r plot_results2, fig.height = 6, fig.width = 11, fig.align = "center"}

cur_df <- df[df$methodW=="lasso",]
p1 <- ggplot(cur_df, aes(x = ldaU, y = U_l0_mean, colour = ldaW)) + geom_point(size = 1) +
  theme_bw() + ylim(0,max(cur_df$U_l0_mean)) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~num_blocks_W, labeller = "label_both")

p2 <- ggplot(cur_df, aes(x = ldaU, y = d_modules, colour = ldaW)) + geom_point(size = 1) +
  theme_bw() + ylim(0,max(cur_df$d_modules)) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~num_blocks_W, labeller = "label_both")

p3 <- ggplot(cur_df, aes(x = ldaU, y = SMAF_W_l0_mean, colour = ldaW)) + geom_point(size = 1) +
  theme_bw() + ylim(0,max(cur_df$SMAF_W_l0_mean)) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~num_blocks_W, labeller = "label_both")

p4 <- ggplot(cur_df, aes(x = ldaU, y = SMAF_Fit, colour = ldaW)) + geom_point(size = 1) +
  theme_bw() + ylim(0,1) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~num_blocks_W, labeller = "label_both")


plot_grid(p1,p2,p3,p4, nrow=2)
```

### U_stability over replicates

#### Read Us
```{r read_Us}
U_list.files <- list.files(U_path, "U*.csv")
# sort U by expid
idx <- lapply(U_list.files, function(x){strsplit(strsplit(x,"expid")[[1]][2],".csv")[[1]][1]})%>% 
                                        unlist() %>% as.numeric() %>% sort(index.return = TRUE) %>% .$ix
U_list.files <- U_list.files[idx]

# read Us into list of df
U_list <- lapply(U_list.files, function(x){read.csv(file.path(U_path,x), check.names = F, row.names = 1) })
```

#### 1. omp_fixedk

```{r subset_df}
df1 <- df[df$methodW == "omp_fixedk",]
df1$cond <- paste0("k=",df1$k," / ldaU=",df1$ldaU)
```

##### combine results_df and U_list into one df
```{r combine_df}
# select necessary results
tmp_df <- dplyr::select(df1, c(replicate,cond,expid))
tmp_df$replicate <- as.factor(tmp_df$replicate) # to use as a group in lpcor
# combine t(U) (modules x genes) and results (tmp_df) as one df. Matched only by their order!
# replicate results for no. of modules, to combine with corresponding t(U)
df_for_lpcor <- lapply(tmp_df$expid, function(expid){
  replicate(U_list[[expid+1]] %>% ncol(), 
            tmp_df[tmp_df$expid == expid,] %>% select(-expid), 
            simplify = FALSE) %>% do.call("rbind",.) %>% cbind(t(U_list[[expid+1]])) 
  }) %>% do.call("rbind",.)
```

##### U_similarity as correlations of gene correlation matrix between U replicates
```{r cor_gcor}
sum_df <- unique(df_for_lpcor$cond) %>% lapply(function(c){
  # subset df for current condition
  cur_df <- df_for_lpcor[df_for_lpcor$cond %in% c,]
  # calc gene correlation matrix (flattened upper triangle as vector) per group ("replicate" in this case)
  # lpcor_tb[[1]] is a vec of grouping ("replicate" in this case)
  # lpcor_tb[[2]][[replicate]]$results[c("Pairs","linear")] is the gene pair and its correlation
  gcor_tb <- cur_df %>% group_by(replicate) %>% lpcor() 
  
  # reshape correlation matrix (as vector) into (gene pairs) x ("replicates") 
  gcor_df <- gcor_tb[[2]] %>% seq_along() %>% 
    lapply(function(i){
                      tmp <- gcor_tb[[2]][[i]]$results[c("Pairs","linear")] 
                      tmp$rep <- i
                      tmp
                      }) %>% do.call("rbind",.) %>% pivot_wider(names_from = rep, values_from = linear)
  
  # calc correlation matrix for "replicates" 
  # (= calc correlation of gene correlation matrices for all combination of replicates)
  # cor_gcor_df is correlation vec for all "replication" pair       
  m <- as.matrix(gcor_df[, -.(Pairs)] )
  cor_gcor_df <- data.frame(correlation = cor(m)[upper.tri(cor(m))], condition = c) 
  return(cor_gcor_df)
})  %>%
  do.call("rbind",.)
```

##### retrieve variables for plotting
```{r add_condition_varible}
# add individual vars
sum_df$k <- sum_df$condition %>% sapply(function(x){ x %>%
  strsplit("/") %>% .[[1]] %>% .[1] %>% 
  strsplit("=") %>% .[[1]] %>% .[2] %>% as.integer() %>% as.factor()
})
sum_df$ldaU <- sum_df$condition %>% sapply(function(x){ x %>%
  strsplit("/") %>% .[[1]] %>% .[2] %>% 
  strsplit("=") %>% .[[1]] %>% .[2] %>% as.numeric() %>% as.factor()
})
```

##### plot
```{r plot_stability, results="asis",  fig.height = 3, fig.width = 5, fig.align = "center"}
ggplot(sum_df, aes(x = ldaU , y = correlation) ) +
  facet_wrap(~k, nrow = 1, labeller = "label_both" ) +
  geom_boxplot() + geom_jitter(colour = "cyan3", size = 1, width = 0.1) + 
  theme_bw() + ylim(0,1) +
  theme(axis.text.x = element_text(angle=45, hjust = 1)) 

```

#### 2. lasso

```{r subset_df2}
df2 <- df[df$methodW == "lasso",]
df2$cond <- paste0("ldaW=",df2$ldaW," / ldaU=",df2$ldaU," / nblocks=",df2$num_blocks_W)
```

##### combine results_df and U_list into one df
```{r combine_df2}
# select necessary results
tmp_df <- dplyr::select(df2, c(replicate,cond,expid))
tmp_df$replicate <- as.factor(tmp_df$replicate) # to use as a group in lpcor
# combine t(U) (modules x genes) and results (tmp_df) as one df. Matched only by their order!
# replicate results for no. of modules, to combine with corresponding t(U)
df_for_lpcor <- lapply(tmp_df$expid, function(expid){
  replicate(U_list[[expid+1]] %>% ncol(), 
            tmp_df[tmp_df$expid == expid,] %>% select(-expid), 
            simplify = FALSE) %>% do.call("rbind",.) %>% cbind(t(U_list[[expid+1]])) 
  }) %>% do.call("rbind",.)
```

##### U_similarity as correlations of gene correlation matrix between U replicates
```{r cor_gcor2}
sum_df <- unique(df_for_lpcor$cond) %>% lapply(function(c){
  # subset df for current condition
  cur_df <- df_for_lpcor[df_for_lpcor$cond %in% c,]
  # calc gene correlation matrix (flattened upper triangle as vector) per group ("replicate" in this case)
  # lpcor_tb[[1]] is a vec of grouping ("replicate" in this case)
  # lpcor_tb[[2]][[replicate]]$results[c("Pairs","linear")] is the gene pair and its correlation
  gcor_tb <- cur_df %>% group_by(replicate) %>% lpcor() 
  
  # reshape correlation matrix (as vector) into (gene pairs) x ("replicates") 
  gcor_df <- gcor_tb[[2]] %>% seq_along() %>% 
    lapply(function(i){
                      tmp <- gcor_tb[[2]][[i]]$results[c("Pairs","linear")] 
                      tmp$rep <- i
                      tmp
                      }) %>% do.call("rbind",.) %>% pivot_wider(names_from = rep, values_from = linear)
  
  # calc correlation matrix for "replicates" 
  # (= calc correlation of gene correlation matrices for all combination of replicates)
  # cor_gcor_df is correlation vec for all "replication" pair       
  m <- as.matrix(gcor_df[, -.(Pairs)] )
  cor_gcor_df <- data.frame(correlation = cor(m)[upper.tri(cor(m))], condition = c) 
  return(cor_gcor_df)
})  %>%
  do.call("rbind",.)
```

##### retrieve variables for plotting
```{r add_condition_varible2}
# add individual vars
sum_df$ldaW <- sum_df$condition %>% sapply(function(x){ x %>%
  strsplit("/") %>% .[[1]] %>% .[1] %>% 
  strsplit("=") %>% .[[1]] %>% .[2] %>% as.numeric() %>% as.factor()
})
sum_df$ldaU <- sum_df$condition %>% sapply(function(x){ x %>%
  strsplit("/") %>% .[[1]] %>% .[2] %>% 
  strsplit("=") %>% .[[1]] %>% .[2] %>% as.numeric() %>% as.factor()
})
sum_df$num_blocks_W <- sum_df$condition %>% sapply(function(x){ x %>%
  strsplit("/") %>% .[[1]] %>% .[3] %>% 
  strsplit("=") %>% .[[1]] %>% .[2] %>% as.numeric() %>% as.factor()
})
```

##### plot
```{r plot_stability2, results="asis",  fig.height = 8, fig.width = 8, fig.align = "center"}
ggplot(sum_df, aes(x = ldaU , y = correlation) ) +
  facet_grid(num_blocks_W~ldaW,  labeller = "label_both" ) +
  geom_boxplot() + geom_jitter(colour = "cyan3", size = 1, width = 0.1) + 
  theme_bw() + ylim(0,1) +
  theme(axis.text.x = element_text(angle=45, hjust = 1)) 



```

