---
title: "9_CISI_actual"
author: "tsuyoshi"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: "hide"
    pandoc_args: [ 
      "--output=9_CISI_actual.html" 
    ]
date: "`r format(Sys.time(), '%d %B, %Y')`"

---

### setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(tidyr)
library(dplyr)
library(ComplexHeatmap)
library(common)
library(colorRamp2)

```


### input data
```{r specify_paths_1}

project.path <- "/mnt/projects/data/publication/9_CISI_actual"
# Specify input paths
genewise.path <- file.path(project.path,"result_genewise.csv")
cisich.path <- file.path(project.path,"result_cisich_corr.csv")
summary.path <- file.path(project.path,"result_summary.csv")
```

```{r}
df_summary <- read.csv(summary.path,  row.names = 'X') 
df_genewise <- read.csv(genewise.path, row.names = 'X')
df_cisich <- read.csv(cisich.path,  row.names = 'X')
```


```{r reshape_df}
# pivot longer for ggplot
df_summary$condition <- rownames(df_summary)
df_summary <-df_summary %>% 
  pivot_longer(c('Gene_average','Gene_minimum','Cell_average'), 
               names_to = "value_type", values_to = "value")

df_genewise$condition <- rownames(df_genewise)
df_genewise <- df_genewise %>% 
  pivot_longer(!condition, names_to = "marker", values_to = "correlation")


df_cisich$cisi_channel <- rownames(df_cisich)
df_cisich <- df_cisich %>%
  pivot_longer(!cisi_channel, names_to = "condition", values_to = "correlation")
```

### plot summary
```{r, results="asis",fig.height = 3, fig.width =10, fig.align = "center"}
ggplot(df_summary, 
       aes(x = condition, y = value, fill = condition)) + 
  facet_wrap(~value_type) +
  geom_bar(stat = "identity") + theme_bw() + xlab(NULL) + 
  theme(axis.text.x=element_blank())
```

### plot all conditions for all proteins
```{r, results="asis",fig.height = 3, fig.width =8, fig.align = "center"}
ggplot(df_genewise, 
       aes(x = reorder(marker, correlation), y = correlation, fill = condition, group = condition)) +
  geom_col(position =  position_dodge2(padding = 0.2),width = 0.7 ) + theme_bw() +
  xlab("marker") + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### plot Actual_Ahat for all proteins
```{r, results="asis",fig.height = 3, fig.width =6, fig.align = "center"}
ggplot(df_genewise[df_genewise$condition == "Actual_Ahat",], 
       aes(x = reorder(marker, correlation, fill = "grey"), y = correlation)) +
  geom_col(width = 0.7 ) + theme_bw() +
  xlab("marker") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```


### read A and Ahat
```{r}
# read A
A_path <- "/mnt/projects/data/publication/9_CISI_actual/A.csv"
A <- read.csv(A_path,row.names = 'X') %>% as.matrix()

# read Ahat
Ahat_path <- "/mnt/projects/data/publication/9_CISI_actual/Ahat.csv"
Ahat <- read.csv(Ahat_path,row.names = 'X') %>% as.matrix()
```

### plot A and Ahat
```{r, results="asis",fig.height = 5, fig.width = 7, fig.align = "center"}


h1 <- Heatmap(t(A), 
              col = colorRamp2(c(0, 0.5, 1), c("navy", "white", "firebrick2")),
              heatmap_legend_param = list(title = "A", direction = "vertical"),
              cluster_columns = F, column_title = 'A: original',
              top_annotation = columnAnnotation(
                corr = anno_barplot(df_cisich[df_cisich$condition=='actual_vs_simulated_A',]$correlation),
                height = unit(3, "cm"),show_annotation_name = FALSE))
h2 <- Heatmap(t(Ahat), 
              col = colorRamp2(c(0, 0.5*max(Ahat), max(Ahat)), c("navy", "white", "firebrick2")),
              heatmap_legend_param = list(title = "Ahat", direction = "vertical"),
              cluster_columns = F, column_title = 'Ahat: reweighted',
              top_annotation = columnAnnotation(
                corr = anno_barplot(df_cisich[df_cisich$condition=='actual_vs_simulated_Ahat',]$correlation, 
                                    axis = F),
                height = unit(3, "cm")))
h1 + h2
# draw(h_list, ht_gap = unit(0.5, "cm"))

```

## single-cell analyses

### load additional libraries
```{r}
library(zellkonverter)
library(SingleCellExperiment)
library(scater)
library(uwot)
library(data.table)
library(RColorBrewer)
library(reshape2)
library(Rphenograph)
library(viridis)
```

### util functions
```{r util_functions}
# conver factor into character in DataFrame (for rbind different sces)
factor2character <- function(DF1){
  DF2 <- DataFrame(lapply(DF1, function(c) {
    if (class(c) == "factor") c <- as.character(c)
    c
  }))
  DF2
}

# get df of cells x reducedDim from sce. Cells are repeated for ngenes times.
get_df <- function(sce, assay = "exprs", long_genes = TRUE) {
    dr <- do.call("cbind", reducedDims(sce))
    foo <- sapply(reducedDims(sce), ncol)
    colnames(dr) <- paste0(rep.int(names(foo), foo), sapply(foo, seq_len))
    df <- data.frame(dr, colData(sce), t(assay(sce, assay)), check.names = FALSE)
    if (long_genes){
      df <- reshape2::melt(df, id.vars = c(colnames(colData(sce)), colnames(dr)))
    }
    df
}

# compute scaled expressions for visualization
scale_sce <- function(sce, assay = "exprs", percent = 0.9999){
  # assay(sce, "exprs") <- asinh(assay(sce, "counts") / 2)
  es <- assay(sce, assay)
  qs <- rowQuantiles(es, probs = c(0.0, percent))
  x <- (es - qs[, 1]) / (qs[, 2] - qs[, 1])
  x[x < 0] <- 0; x[x > 1] <- 1
  return(x)
}

# plot reduced dimensions with facets
plot_dr <- function(sce, assay = "scaled", dr_type = "TSNE", facet = "variable"){
    plot <- data.table(get_df(sce, assay = assay)) %>%
    ggplot(aes(get(paste0(dr_type,1)), get(paste0(dr_type,2)), col = value)) +
    geom_point(alpha=0.99, size=0.1) + 
    facet_wrap(facet, ncol = 5) +
    scale_color_gradientn(colours=rev(brewer.pal(11, 'Spectral')))+
    theme_void() + theme(aspect.ratio = 1, strip.text.x = element_text(size = 12))
    plot
}

# plot reduced dimensions with colors
plot_dr2 <- function(sce, assay = "scaled", dr_type = "TSNE", colourby = "tissue", legendPsize = 5){
    plot <- data.table(get_df(sce, assay = assay, long_genes = FALSE)) %>%
    ggplot(aes(get(paste0(dr_type,1)), get(paste0(dr_type,2)), col = get(colourby))) +
    geom_point(alpha=0.99, size=0.1) + 
    #scale_color_viridis(discrete = TRUE)+
    theme_void() + theme(aspect.ratio = 1,strip.text.x = element_text(size = 8))+ 
    guides(colour = guide_legend(override.aes = list(size=legendPsize)))
    plot
}

``` 

### set input
```{r specify_paths_2}
# Specify input paths
sce_gt.path <- file.path(project.path,"sce/sce_gt.h5ad")
sce_dc_Ahat.path <- file.path(project.path,"sce/sce_dc_Actual_Ahat.h5ad")

sce_gt.path <- file.path("/mnt/projects/data/11_realCISI_5tissues/sce_gt.h5ad")
sce_dc_Ahat.path <- file.path("/mnt/projects/data/14_recalc_X/sce_dc.h5ad")

# also read original sce for panCK signals
sce_ori.path <- file.path("/mnt/projects/data/0_preprocess_th186/processed","filtered_sce.h5ad")

```

### input sces
```{r}
# read h5ad to sce
sce_gt <- readH5AD(sce_gt.path) 
sce_dc_Ahat <- readH5AD(sce_dc_Ahat.path)

# combine sces into list
SCE <- list(sce_gt,sce_dc_Ahat)# list(sce_gt,sce_dc,sce_dc_Ahat)
names(SCE) <- c("ground_truth", "decompressed_reweightedA" )

# ori is for panCK data
sce_ori <- readH5AD(sce_ori.path)
```
```{r}
# prepare panck data
panck <- sce_ori["panCK",match(colnames(sce_gt),colnames(sce_ori))]
colData(panck) <- factor2character(colData(panck))
```


```{r}
# convert factor in coldata into character and rbind panCK data
SCE <- lapply(SCE, function(sce) {
  colData(sce) <- factor2character(colData(sce) )
  sce <- rbind(sce, panck)
})

```



### transform and scale counts for UMAP and visualisation
```{r add_exprs_scaled}
SCE <- lapply(SCE, function(sce) {
              assay(sce, "exprs") <- asinh(counts(sce)/1)
              assay(sce, "scaled") <- scale_sce(sce)
              sce
            })

```


### run UMAP
```{r runUMAP}
SCE <- lapply(SCE, function(sce){
  set.seed(230905)
  sce <- runUMAP(sce, exprs_values="exprs")
  sce
})
  
# sce <- runTSNE(sce, exprs_values="exprs")
```

### run Phenograph
```{r runPhenograph}
SCE <- lapply(SCE, function(sce){
  set.seed(230907) # need to seed every calculation 
  pg_res <- Rphenograph(data = t(assay(sce,"exprs")), k = 15)
  sce$PG <- factor(pg_res[[2]]$membership)
  sce
})

```

### plot clusters
```{r}
HM <- lapply(SCE, function(sce){
    summary_dat = data.table(get_df(sce,assay = "scaled"))[,list(
      median_val = median(value), mean_val = mean(value), cell_cluster=.N), by=.(variable,PG)]
  #median values per cluster
  hm_dat = dcast.data.table(data =summary_dat, formula = 'PG ~ variable', value.var = 'median_val')#'median_val'
  group_levels = factor(hm_dat$PG)
  trownames = hm_dat$PG
  # Convert to a matrix
  hm_dat = as.matrix(hm_dat[,-1,with=F])
  row.names(hm_dat) = trownames
  hm_dat
})
```

```{r warning = FALSE, fig.width=6, fig.height=7}
Heatmap(HM$ground_truth, name = "ground_truth", 
        col = viridis::magma(11), rect_gp = gpar(col = "white", lwd = 1),
        right_annotation = rowAnnotation(
          Ncells = anno_barplot(as.vector(table(as.integer(SCE$ground_truth$PG))))))


Heatmap(HM$decompressed_reweightedA, name = "decompressed\nreweightedA", 
        col = viridis::magma(11), rect_gp = gpar(col = "white", lwd = 1),
        right_annotation = rowAnnotation(
          Ncells = anno_barplot(as.vector(table(as.integer(SCE$decompressed_reweightedA$PG))))))
```

### annotate clusters
```{r}
# annotation for GT
anno_1 <- list(B = c(29,27,33),
               TandB = c(40),
               Th = c(34,26,20,4),
               Tc = c(11),
               plasma = c(23,18,14),
               pDC = c(21),
               migDC = c(30),
               Macrophage = c(24,15,1),
               Neutrophil = c(17,16,12),
               NK = c(19),
               Endothelial = c(13,25,7,6),
               Fibroblast = c(9,2,3,8),
               Tumor = c(32,36,39,38,37,35,31,22),
               Ki67 = c(10),
               Unidentified = c(5,28))
# initialize annotation
SCE$ground_truth$annotation <- "not_annotated"

# assign to sce$annotation
for (i in seq_along(anno_1)){
  SCE$ground_truth[,SCE$ground_truth$PG %in% anno_1[[i]]]$annotation <- names(anno_1)[[i]]
}
```

```{r}
# annotation for decompressed_reweightedA
anno_2 <- list(B = c(24,23,27,19),
               TandB = c(30),
               Th = c(12,25,35),
               Tc = c(20,3),
               plasma = c(21,22),
               pDC = c(17),
               migDC = c(34),
               Macrophage = c(13,2,8),
               Neutrophil = c(15),
               NK = c(28),
               Endothelial = c(6),
               Fibroblast = c(4,7,5),
               Tumor = c(31,32,29,9,33,11),
               Ki67 = c(16,26),
               Unidentified = c(18,1,14,10))

# initialize annotation
SCE$decompressed_reweightedA$annotation <- "not_annotated"

# assign to sce$annotation
for (i in seq_along(anno_2)){
  SCE$decompressed_reweightedA[,SCE$decompressed_reweightedA$PG %in% anno_2[[i]]]$annotation <- names(anno_2)[[i]]
}
```
```{r prep_annotation}
# prepare annotation of PG of Gt to its celltype annotation
GT_PG_to_GT_Anno <- 1:length(unique(SCE$ground_truth$PG))
for (i in seq_along(anno_1)){
  GT_PG_to_GT_Anno[match(anno_1[[i]], GT_PG_to_GT_Anno)] <- names(anno_1)[[i]]
}

DC_PG_to_DC_Anno <- 1:length(unique(SCE$decompressed_reweightedA$PG))
for (i in seq_along(anno_2)){
  DC_PG_to_DC_Anno[match(anno_2[[i]], DC_PG_to_DC_Anno)] <- names(anno_2)[[i]]
}

# prepare annotation for ncells
GT_PG_Ncells_Anno = row_anno_barplot(as.vector(table(as.integer(SCE$ground_truth$PG))))
DC_PG_Ncells_Anno = row_anno_barplot(as.vector(table(as.integer(SCE$decompressed_reweightedA$PG))))
```

### plot annotaated celltypes on heatmap of median expression per cluster
```{r, fig.width=8, fig.height=8}
Heatmap(HM$ground_truth, name = "ground_truth", 
              col = viridis::viridis(11), rect_gp = gpar(col = "white", lwd = 1), 
              right_annotation = rowAnnotation(GT_anno = anno_text(GT_PG_to_GT_Anno),
                                               Ncells = GT_PG_Ncells_Anno))
Heatmap(HM$decompressed_reweightedA, name = "decompressed\nreweightedA", 
              col = viridis::viridis(11), rect_gp = gpar(col = "white", lwd = 1), 
              right_annotation = rowAnnotation(DC_anno = anno_text(DC_PG_to_DC_Anno),
                                               Ncells = DC_PG_Ncells_Anno))
```

### plot annotaated celltypes on umap
```{r, fig.width=8, fig.height=8}
plot_dr2(SCE$ground_truth, assay = "scaled", dr_type = "UMAP", colourby = "annotation")
plot_dr2(SCE$decompressed_reweightedA, assay = "scaled", dr_type = "UMAP", colourby = "annotation")
```


### heatmaps of matching ground truth cell type and decompressed cell type
```{r}
# table with row of GT celltype and col of DC celltype
mat1 <- table(SCE$ground_truth$annotation, SCE$decompressed_reweightedA$annotation) %>%
  prop.table(margin = 1)

mat2 <- table(SCE$ground_truth$annotation, SCE$decompressed_reweightedA$annotation) %>%
  prop.table(margin = 2)

```
```{r, fig.width=7, fig.height=7}
plot_colwise_Heatmap <-
  function(mat2, legend_title ,row_title, col_title){
    hs <- NULL
    for (i in 1:ncol(mat2)){
      if (i == 1){
        h <- Heatmap(mat2[,i],  name = colnames(mat2)[i],
            cluster_columns = F, cluster_rows = F, border_gp = gpar(col = "black", lty = 1),
            width = 1*unit(5, "mm"), height = ncol(mat2)*unit(5, "mm"),
            col = colorRamp2(c(0, 1), c("white", "navy")),show_heatmap_legend = (i==1),
            heatmap_legend_param = list(title = legend_title),
            row_title = row_title, row_title_side = "left")
      } else if (i == as.integer(ncol(mat2)/2)){
        h <- Heatmap(mat2[,i],  name = colnames(mat2)[i],
            cluster_columns = F, cluster_rows = F, border_gp = gpar(col = "black", lty = 1),
            width = 1*unit(5, "mm"), height = ncol(mat2)*unit(5, "mm"),
            col = colorRamp2(c(0, 1), c("white", "navy")),show_heatmap_legend = (i==1),
            heatmap_legend_param = list(title = legend_title),
            column_title = col_title, column_title_side = "bottom")
      }else{
        h <- Heatmap(mat2[,i],  name = colnames(mat2)[i],
            cluster_columns = F, cluster_rows = F, border_gp = gpar(col = "black", lty = 1),
            width = 1*unit(5, "mm"), height = ncol(mat2)*unit(5, "mm"),
            col = colorRamp2(c(0, 1), c("white", "navy")),show_heatmap_legend = (i==1),
            heatmap_legend_param = list(title = legend_title)
            )
      }
      hs <- hs + h
    }
  return(hs)
  }
  
plot_rowwise_Heatmap <-
  function(mat1, legend_title ,row_title, col_title){
    hs <- NULL
    for (i in 1:nrow(mat1)){
      if (i == 1){
        h <- Heatmap(mat1[i,],  name = rownames(mat1)[i],
            cluster_columns = F, cluster_rows = F, border_gp = gpar(col = "black", lty = 1),
            width = 1*unit(5, "mm"), height = nrow(mat1)*unit(5, "mm"),
            col = colorRamp2(c(0, 1), c("white", "navy")),show_heatmap_legend = (i==1),
            heatmap_legend_param = list(title = legend_title),
            row_title = row_title, row_title_side = "left")
      } else if (i == as.integer(nrow(mat1)/2)){
        h <- Heatmap(mat1[i,],  name = rownames(mat1)[i],
            cluster_columns = F, cluster_rows = F, border_gp = gpar(col = "black", lty = 1),
            width = 1*unit(5, "mm"), height = nrow(mat1)*unit(5, "mm"),
            col = colorRamp2(c(0, 1), c("white", "navy")),show_heatmap_legend = (i==1),
            heatmap_legend_param = list(title = legend_title),
            column_title = col_title, column_title_side = "bottom")
      }else{
        h <- Heatmap(mat1[i,],  name = rownames(mat1)[i],
            cluster_columns = F, cluster_rows = F, border_gp = gpar(col = "black", lty = 1),
            width = 1*unit(5, "mm"), height = nrow(mat1)*unit(5, "mm"),
            col = colorRamp2(c(0, 1), c("white", "navy")),show_heatmap_legend = (i==1),
            heatmap_legend_param = list(title = legend_title)
            )
      }
      hs <- hs + h
    }
  return(hs)
  }


plot_colwise_Heatmap(mat2,"proportion\ndecompressed","Ground_truth","Decompressed")
plot_rowwise_Heatmap(mat1, "proportion\ngeound_truth","Decompressed","Ground_truth")
```


### calculate F1 Recall Precision
```{r}
recall <- diag(mat1)
precision <- diag(mat2)
f1 <- 2*recall*precision/(recall + precision)

df <- data.frame(recall = recall, precision = precision, f1 = f1)
df$annotation <- rownames(df)
df
```

```{r, results="asis",fig.height = 3, fig.width =6, fig.align = "center"}
ggplot(df %>% pivot_longer(!annotation, names_to = "value_type", values_to = "score"),
       aes(x= annotation, y = score, fill = value_type))+  
  geom_col(width = 0.7, position = position_dodge2() ) + theme_bw() +
  xlab("annotation") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

cat("Mean F1 score: ",mean(df$f1),"\n")
cat("Mean Recall: ",mean(df$recall),"\n")
cat("Mean Precision: ",mean(df$precision),"\n")
```





